Modelling Prehistoric Iconographic Compositions. The \proglang{R} package \pkg{decorr}

### 2020-12-12 article_rvTH17.Rnw
new version

<<ed, results='asis', echo=TRUE, fig.cap="First edge of the Torrejon Rubio 1 decoration, the proximity between the sword (GU 6) and the fibula (GU 5)">>=
caption <- paste0("First edge of the Torrejon Rubio 1 decoration, ",
                  "the proximity between the sword (node 6) and the fibula (node 5)")
eds.df <- read_eds(site = sit, decor = dec, dev = ".tsv",
                   doss = system.file("extdata", package = "decorr"))
print(xtable::xtable(eds.df[1,],
                     caption = caption,
                     label = "tab:ed",
                     size = 7),
      table.placement = "!htbp")
@
% \end{figure}

@@

<<echo=TRUE, message=FALSE, warning=FALSE, fig.pos = '!htbp', fig.align="center", fig.width=8, fig.height=4.2, fig.cap="\\label{fig:ceds}comparisons between 1 (Cerro Muriano 1) and 4 (Zarza de Montsanchez) decorations">>=
par(mar=c(0,0,0,0))
g.compar <- list_compar(lgrph,"type")
eds_compar <- plot_compar(g.compar, c(1,4),
                          focus = "edges",
                          doss = system.file("extdata", package = "decorr"))
plot(magick::image_trim(magick::image_read(eds_compar)))
@

@@

As said, since the precise location of the nodes is usualy not registred, the most commonly used method in statistical analysis on prehistorical iconography is the presence of common nodes. In order to demonstrate the first insight of a graph-based analysis of the decorations, we will compare two classifications, the first one based on the "classic" presence of common nodes, the second one based on the presence of common edges.

\begin{table}[!htbp]
\centering
\begin{minipage}{.3\textwidth}
\centering
\textbf{Common nodes}\raisebox{-1ex}{}\\
<<echo=FALSE,results='asis' >>=
library("xtable")
# library("MASS");library(decorr);library(xtable);library(magick)
# imgs <- read.table(system.file("extdata", "imgs.tsv", package = "decorr"),
#                    sep="\t",stringsAsFactors = FALSE)
# nodes <- read.table(system.file("extdata", "nodes.tsv", package = "decorr"),
#                     sep="\t",stringsAsFactors = FALSE)
# edges <- read.table(system.file("extdata", "edges.tsv", package = "decorr"),
#                     sep="\t",stringsAsFactors = FALSE)
# lgrph <- list_dec(imgs,nodes,edges,"type")
df.same_edges <- same_elements(lgrph, "type", "edges")
df.same_nodes<- same_elements(lgrph, "type", "nodes")
# library("xtable")
print(xtable::xtable(df.same_nodes,
                     size = 8,
                     digits = c(0)),
      floating=FALSE,
      include.rownames = TRUE)
# print(xtable(df.same_nodes, digits=c(0)),
#       floating=FALSE,
#       hline.after=NULL)
@
\end{minipage}
\begin{minipage}{.3\textwidth}
\centering
\textbf{Common edges}\raisebox{-1ex}{}\\
<<echo=FALSE,results='asis'>>=
print(xtable::xtable(df.same_edges,
                     size = 8,
                     digits = c(0)),
      floating=FALSE,
      include.rownames = TRUE)
# print(xtable(df.same_edges, digits=c(0)),
#       floating=FALSE,
#       hline.after=NULL)
@
\end{minipage}
\caption{Tables showing the number of common nodes and common edges between each pair of decorations}
\end{table}

Once the heatmap matrices calculated, the native \code{dist()} and \code{hclust()} functions \citep{R19} are calculated from the inverse matrices with the function \code{inv()} of the \pkg{matlib} package \citep{Friendly20}

\begin{figure}[!htbp]
<<echo=TRUE, warning=FALSE>>=
library("matlib")
par(mfrow=c(1,2))
dist.nodes  <- dist(inv(as.matrix(df.same_nodes)))
dist.edges  <- dist(inv(as.matrix(df.same_edges)))
plot(hclust(dist.nodes), hang = -1, main = "common nodes")
plot(hclust(dist.edges), hang = -1, main = "common edges")
@
\end{figure}

Results of classifications show that for both common nodes and common edges, the most different decorations are \code{1} and \code{4}. These two decorations share four (4) common nodes and, as previously seen, only two (2) common edges.
In any cases decorations \code{2} and \code{3} are closer to decoration \code{4} than to decoration \code{1}, but their classifications changes depending on counting of common nodes or common edges. While decorations \code{2} and \code{4} have 7 common edges and 7 common nodes, plotting the comparisons for \code{3} and \code{4}, helps to understand the differences between the two classifications.

<<echo=TRUE, message=FALSE, warning=FALSE, fig.pos = '!htbp', fig.align="center", fig.cap="\\label{fig:figs}Decoration comparisons between 3 and 4">>=
par(mar=c(0,0,0,0))
par(mfrow=c(2,1))
g.compar <- list_compar(lgrph,"type")
nds_compar.3.4 <- plot_compar(g.compar, c(3,4),
                              focus = "nodes",
                              doss = system.file("extdata", package = "decorr"))
plot(magick::image_read(nds_compar.3.4))
# g.compar <- list_eds_compar(lgrph,"type")
eds_compar.3.4 <- plot_compar(g.compar, c(3,4),
                              focus = "edges",
                              doss = system.file("extdata", package = "decorr"))
plot(magick::image_read(eds_compar.3.4))
@

Decorations \code{3} and \code{4} share four (4) common nodes (\code{bouclier, epee, lance, miroir}) but  these nodes have different spatial organisations with only one common edge (\code{bouclier -=- lance}). At the opposite, decorations \code{2} and \code{4} show more properly the same compositions.

>>>>

<<ed, results='asis', echo=TRUE >>=
caption <- paste0("First edge of the Torrejon Rubio 1 decoration, ",
                  "the proximity between the sword (node 6) ",
                  "and the fibula (node 5)")
eds.df <- read_eds(site = sit, decor = dec, dev = ".tsv",
                   doss = system.file("extdata", package = "decorr"))
print(xtable::xtable(eds.df[1,],
                     caption = caption,
                     label = "tab:ed",
                     size = 7),
      table.placement = "!htbp")
@
% \end{figure}

@@

<<echo=TRUE, message=FALSE, warning=FALSE, fig.pos = '!htbp', fig.align="center", fig.width=8, fig.height=4.2, fig.cap="\\label{fig:ceds}Comparisons between decoration 1 (Cerro Muriano 1) and decoration 4 (Zarza de Montsanchez)">>=
par(mar=c(0,0,0,0))
g.compar <- list_compar(lgrph, "type")
eds_compar <- plot_compar(g.compar, c(1, 4),
                          focus = "edges",
                          doss = system.file("extdata", package = "decorr"))
plot(magick::image_trim(magick::image_read(eds_compar)))
@

@@

In order to demonstrate the main insight of a graph-based analysis of the decorations, we will compare two classifications, the first one based on the "classic" presence of common nodes -- since the precise location of the nodes is usualy not registred in the studies --, the second one based on the presence of common edges.

At first, the Late Bronze Age phase of Ibahernando stelae must be selected (ie, the Roman latin writing must be removed). After the selection of the graph elements of Ibahernando, the connected component of the node 2, the lance (spear), is selected with \code{contemp_nds()} and the list of graphs (\code{lgrph}) is updated

<<results='asis', echo=TRUE, fig.height= 5>>=
dataDir <- system.file("extdata", package = "decorr")
# Read a decoration
nds.df <- read_nds(site = "Ibahernando",
                   decor = "Ibahernando",
                   doss = dataDir)

eds.df <- read_eds(site = "Ibahernando",
                   decor = "Ibahernando",
                   doss = dataDir)
# Extract the graph contemporaneous to the node 2
l_dec_df <- contemp_nds(nds.df, eds.df, selected.nd = 2)
# remove first co
l_dec_df[[1]]$site <- l_dec_df[[1]]$decor <- NULL
l_dec_df[[2]]$site <- l_dec_df[[2]]$decor <- NULL
g <- graph_from_data_frame(l_dec_df[[2]], vertices = l_dec_df[[1]],
                           directed = FALSE)
# re-add the site and decor columns 
# with Ibahernando values and the graph name
g$site <- g$decor <- "Ibahernando"
g$name <- 5
# overwrite the decor in the list
lgrph[[5]] <- g
@

Then, \code{same_elements()} is done on nodes and edges (Table \ref{tab:common.elts})

<<results='asis', echo=TRUE>>=
df.same_edges <- same_elements(lgrph, "type", "edges")
df.same_nodes<- same_elements(lgrph, "type", "nodes")
@

\begin{table}[!htbp]
% \begin{table}[H]
\centering
\begin{minipage}{.3\textwidth}
\centering
\textbf{Common nodes}\raisebox{-1ex}{}\\
<<echo=FALSE,results='asis' >>=
print(xtable::xtable(df.same_nodes,
                     size = 8,
                     digits = c(0)),
      floating=FALSE,
      include.rownames = TRUE)
@
\end{minipage}
\begin{minipage}{.3\textwidth}
\centering
\textbf{Common edges}\raisebox{-1ex}{}\\
<<echo=FALSE,results='asis'>>=
print(xtable::xtable(df.same_edges,
                     size = 8,
                     digits = c(0)),
      floating=FALSE,
      include.rownames = TRUE)
@
\end{minipage}
\caption{Tables showing the number of common nodes and common edges between each pair of decorations}
\label{tab:common.elts}
\end{table}

% the inverse values of common nodes and common edges are calculated with the function \code{inv()} from the package \pkg{matlib} \citep{Friendly20}, and then 
Once the heatmap matrices calculated, the distance matrices are calculated with the native \code{dist()} function \citep{R19} (Table \ref{tab:dist.elts}).

\begin{table}[!htbp]
% \begin{table}[H]
\centering
\begin{minipage}{.4\textwidth}
\centering
\textbf{Distance matrix on nodes}\raisebox{-1ex}{}\\
<<echo=FALSE,results='asis', warning=F >>=
library("matlib")
dist.nodes  <- dist(as.matrix(df.same_nodes))
mat.nodes <- as.matrix(dist.nodes)
mat.nodes[lower.tri(mat.nodes, diag = TRUE)] <- NA
print(xtable::xtable(mat.nodes,
                     size = 8,
                     digits = c(2)),
      floating=FALSE,
      include.rownames = TRUE)
@
\end{minipage}
\begin{minipage}{.4\textwidth}
\centering
\textbf{Distance matrix on edges}\raisebox{-1ex}{}\\
<<echo=FALSE,results='asis'>>=
dist.edges <- dist(as.matrix(df.same_edges))
mat.edges <- as.matrix(dist.edges)
mat.edges[lower.tri(mat.edges, diag = TRUE)] <- NA
print(xtable::xtable(mat.edges,
                     size = 8,
                     digits = c(2)),
      floating=FALSE,
      include.rownames = TRUE)
@
\end{minipage}
\caption{Distance matrices on nodes and common edges between each pair of decorations}
\label{tab:dist.elts}
\end{table}

For both nodes and edges, the results show that the most important differences are between the decoration 1 and the decoration 2: for nodes, dist = \Sexpr{round(max(mat.nodes, na.rm = T), 2)} ; for edges, dist =  \Sexpr{round(max(mat.edges, na.rm = T), 2)}. At the opposite, the decorations showing the more similar compositions are decoration 3 and decoration 5: for nodes, dist = \Sexpr{round(min(mat.nodes, na.rm = T), 2)} ; for edges, dist =  \Sexpr{round(min(mat.edges, na.rm = T), 2)}.
A hierachical clustering (HC) of the distance matrices, with the native function \code{hclust()} \citep{R19}, summarize these similarites/dissimilarities (Figure \ref{fig:hclust})

\begin{figure}[H]
% \begin{figure}[!htbp]
<<echo=TRUE, warning=FALSE, fig.height=5>>=
par(mfrow=c(1,2))
# dist.nodes  <- dist(inv(as.matrix(df.same_nodes)))
# dist.edges  <- dist(inv(as.matrix(df.same_edges)))
# plot(hclust(dist.nodes), hang = -1, main = "common nodes")
# plot(hclust(dist.edges), hang = -1, main = "common edges")
plot(hclust(dist.nodes), main = "common nodes")
plot(hclust(dist.edges), main = "common edges")
@
\label{fig:hclust}
\end{figure}

These latter decorations (decoration 1 and decoration 2) can be plotted with \code{plot_compar()} to vizualise the common elements (Figure \ref{fig:dec1.dec2}).

\begin{figure}[H]
<<echo=TRUE, message=FALSE, warning=FALSE, fig.pos = '!htbp', fig.align="center", fig.cap="\\label{fig:dec1.dec2}Decoration comparisons between decoration 1 and decoration 2">>=
par(mar = c(0, 0, 0, 0))
par(mfrow = c(2, 1))
lgrph <- list_dec(imgs, nodes, edges)
g.compar <- list_compar(lgrph, "type")
nds_compar <- plot_compar(g.compar, c(3, 5),
                          focus = "nodes",
                          nd.size = c(0.5, 1.5),
                          ed.width = c(1, 2.5),
                          lbl.size = .7,
                          doss = system.file("extdata", 
                                             package = "decorr"))
plot(magick::image_read(nds_compar))
eds_compar <- plot_compar(g.compar, c(3, 5),
                          focus = "edges",
                          nd.size = c(0.5, 1.5),
                          ed.width = c(1, 2.5),
                          lbl.size = .7,
                          doss = system.file("extdata", 
                                             package = "decorr"))
plot(magick::image_read(eds_compar))
@
\label{fig:dec1.dec2}
\end{figure}

Even the classifications on nodes and edges give the same results,
